Performance benefits of Intel Optane DC storage for the parallel processing of large neuroimaging data

### 2019-10-22 paper.tex
minor fix
updates to text and figure generation fixes

Due to the size and uniqueness of the \bigbrain, standardized processing pipelines
have yet to be developed for it. In order to quantify the effects of data location
on such a dataset, we have implemented a naive synthetic application that takes 
the image, split into blocks, and increments all the voxels within each block, in parallel (Algorithm~\ref{alg:incrementation}).
This application
enables us to read and write to different storage devices and ensure that written
data could not have been previously cache
in-memory, by ensuring that read and written data are not the same. This application
will be parallelized using both GNU Parallel and Apache Spark 2.4.3 and implemented in Python 3.6 \tristan{In a subsequent 
version you could give more details about implementations, including containerization, etc}. While the GNU Parallel application
is currently limited to 1 incrementation step, the Apache Spark implementation allows us to increase
the number of incrementation steps by a variable amount, thereby increasing the amount of intermediate
data generated.

@@

We plan to execute this pipeline on both the $\SI{76}{\gibi\byte}$ 40~$\mu$m 
\bigbrain split into 125 $\SI{617}{\mega\byte}$ blocks and the $\SI{603}{\gibi\byte}$
\bigbrain at 20~$\mu$m split into 1000 $\SI{617}{\mebi\byte}$ blocks.

At this moment, we have executed the application using GNU Parallel for parallelization
using 25 and 96 processes on the 125 40$\mu$m \bigbrain blocks. Data was read and written to either
tmpfs, Optane (App Direct only), local disk and Isilon. In all configurations (25 and 96 processes)
only 1 iteration was performed. At 25 processes, a sleep delay of 120 seconds was added at the
incrementation step, which would emulate tasks of longer duration and ensure that we satisfy Equation~\ref{eq:page-cache-inequality} for all devices.
At 96 processes with 0 sleep delay, we estimate being above the inequality for local disk and Isilon.
All experiments were repeated 5x in each environment (App Direct and Memory Mode)

@@

Time in this application was obtained using Linux's \texttt{time(1)} application.
\texttt{Real} was used to measure CPU time, where \texttt{Sys} was used to measure
I/O time.
\section{Results}

\subsection{\bigbrain incrementation}
\subsubsection{25 Processes}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-25cpus-120delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{Makespan of \bigbrain Incrementation using 25 processes with a sleep delay of 120s on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:makespan-25cpus}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-25cpus-120delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{Total read/increment/write breakdown of \bigbrain Incrementation using 25 processes with a sleep delay of 120s on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:stacked-25cpus}
\end{figure}
\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1567709282-tmpfsAD_1it_25cpus_120delay_withpc-2.pdf}
    \caption{tmpfs}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1567705458-optaneAD_1it_25cpus_120delay_withpc-2.pdf}
    \caption{Optane}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1567704324-localAD_1it_25cpus_120delay_withpc-2.pdf}
    \caption{local disk}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1567707833-isilonAD_1it_25cpus_120delay_withpc-2.pdf}
    \caption{Isilon}\label{fig:gantt25isilon}
\end{subfigure}
\caption{Gantt charts for each storage device (App Direct Mode) processing 125 blocks of the 40$\mu$m BigBrain using 25 processes and a sleep delay of 120s}\label{fig:gantt25}
\end{figure*}

As shown in Figure~\ref{fig:makespan-25cpus}, despite being under the inequality, the
performance still varies according to selected storage device. As expected, Optane performs
slightly better than local disk and significantly better than Isilon. Tmpfs outperforms all
devices, with only slightly better performance than Optane and local disk, in both App Direct Mode and Memory Mode.
There is no apparent difference between the two modes on the given devices.
\tristan{You should explain why Optane is not reported in memory mode.}

When observing the total task breakdown (Figure~\ref{fig:stacked-25cpus}), it is
apparent that the differences between the makespans arise from difference in incrementation
time between all devices except Isilon. In addition, while tmpfs and local disk have comparable write times, Optane
has a more significant write time difference. Whereas, Isilon has more significant read and
write time compared to all other devices, it's incrementation time is comparable to that
of local disk. Further inspection of each storage device's Gantt chart (see Figure~\ref{fig:gantt25})
show that increased duration of tasks stagger the execution of each individual tasks. This, in turn,
results in better performance towards the end of the execution, where tasks are significantly more staggered.
Only tmpfs appears to be able to execute each task efficiently with little staggering.
\subsubsection{96 Processes}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-96cpus-0delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{Makespan of \bigbrain Incrementation using 96 processes on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:makespan-96cpus}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-96cpus-0delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{total read/increment/write breakdown of \bigbrain incrementation using 96 processes on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:stacked-96cpus}
\end{figure}

@@

\subsection{BIDS App Example}
\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bids-app-example/25cores.pdf}
    \caption{25 processes}\label{fig:bm25}
\end{subfigure}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bids-app-example/96cores.pdf}
    \caption{96 processes}\label{fig:bm96}
\end{subfigure}
\caption{Makespan of BIDS App Example 25 and 96 processes on all storage devices. 3 repetitions were performed}
\end{figure*}

>>>>

Due to the size and uniqueness of the \bigbrain, standardized processing pipelines
have yet to be developed for it. In order to quantify the effects of data location
on such a dataset, we have implemented a naive synthetic application that takes 
the image, split into blocks, and increments all the voxels within each block, in parallel (Algorithm~\ref{alg:incrementation}).
This application
enables us to read and write to different storage devices and ensure that written
data could not have been previously cache
in-memory, by ensuring that read and written data are not the same. This application
was parallelized us in two different ways: using both GNU Parallel and Apache Spark 2.4.3 (PySpark). 
All task code was implemented in Python 3.6 \tristan{In a subsequent 
version you could give more details about implementations, including containerization, etc}. While the GNU Parallel application
is currently limited to 1 incrementation step, the Apache Spark implementation allows us to increase
the number of incrementation steps by a variable amount, thereby increasing the amount of intermediate
data generated.

@@

We have executed this pipeline on both the $\SI{76}{\gibi\byte}$ 40~$\mu$m 
\bigbrain split into 125 $\SI{617}{\mega\byte}$ blocks and the $\SI{603}{\gibi\byte}$
\bigbrain at 20~$\mu$m split into 1000 $\SI{617}{\mebi\byte}$ blocks.

For the 40~$\mu$m \bigbrain, we have executed the application using GNU Parallel for parallelization
using 25 and 96 processes on the 125 40$\mu$m \bigbrain blocks. Data was read and written to either
tmpfs, Optane (App Direct only), local disk and Isilon. In all configurations (25 and 96 processes)
only 1 iteration was performed. At 25 processes, a sleep delay of 120 seconds was added at the
incrementation step, which would emulate tasks of longer duration and ensure that we satisfy Equation~\ref{eq:page-cache-inequality} for all devices.
At 96 processes with 0 sleep delay, we estimate being above the inequality for local disk and Isilon.
All experiments were repeated 5x in each environment (App Direct and Memory Mode).

For the 20~$\mu$m \bigbrain, we used the same application, however in this case, it was executed
using both Apache Spark and GNU Parallel. The configuration was also generally the same, having 
experiments using both 25 and 96 cpus. As the previous experiments would have already shown how
page cache availability for the dirty writes can compare with Optane, we did not include
a task duration delay. Rather, we just measure the performance differences between the various
storage devices, in addition to comparing how each device scales by modifying the cluster size.

Since storage was limited on tmpfs and local disk and was not large enough to process the entire
20~$\mu$m \bigbrain, we had to emulate the reading and writing to these devices. We achieved this
by copying only 70 of the 1000 blocks onto the storage devices, but used 1000 tasks to read, increment and
write these 70 blocks.

@@

Time in this application was obtained using Linux's \texttt{time(1)} application.
\texttt{Real + Sys} was used to measure CPU time,
whereas \texttt{User - (Real + Sys)} was used to measure
I/O time.
\section{Results}

\subsection{40~$\mu$m \bigbrain incrementation}
\subsubsection{25 Processes}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-25cpus-120delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{Makespan of \bigbrain Incrementation using 25 processes with a sleep delay of 120s on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:makespan-25cpus}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-25cpus-120delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{Total read/increment/write breakdown of \bigbrain Incrementation using 25 processes with a sleep delay of 120s on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:stacked-25cpus}
\end{figure}
\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1567709282-tmpfsAD_1it_25cpus_120delay_withpc-2.pdf}
    \caption{tmpfs}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1567705458-optaneAD_1it_25cpus_120delay_withpc-2.pdf}
    \caption{Optane}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1567704324-localAD_1it_25cpus_120delay_withpc-2.pdf}
    \caption{local disk}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1567707833-isilonAD_1it_25cpus_120delay_withpc-2.pdf}
    \caption{Isilon}\label{fig:gantt25isilon}
\end{subfigure}
\caption{Gantt charts for each storage device (App Direct Mode) processing 125 blocks of the 40$\mu$m BigBrain using 25 processes and a sleep delay of 120s}\label{fig:gantt25}
\end{figure*}

As shown in Figure~\ref{fig:makespan-25cpus}, despite being under the inequality, the
performance still varies according to selected storage device. As expected, Optane performs
slightly better than local disk and significantly better than Isilon. Tmpfs outperforms all
devices, with only slightly better performance than Optane and local disk, in both App Direct Mode and Memory Mode.
There is no apparent difference between the two modes on the given devices.

When observing the total task breakdown (Figure~\ref{fig:stacked-25cpus}), it is
apparent that the differences between the makespans arise from difference in incrementation
time between all devices except Isilon. In addition, while tmpfs and local disk have comparable write times, Optane
has a more significant write time difference. Whereas, Isilon has more significant read and
write time compared to all other devices, it's incrementation time is comparable to that
of local disk. Further inspection of each storage device's Gantt chart (see Figure~\ref{fig:gantt25})
show that increased duration of tasks stagger the execution of each individual tasks. This, in turn,
results in better performance towards the end of the execution, where tasks are significantly more staggered.
Only tmpfs appears to be able to execute each task efficiently with little staggering.
\subsubsection{96 Processes}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-real-96cpus-0delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{Makespan of \bigbrain Incrementation using 96 processes on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:makespan-96cpus}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-real-96cpus-0delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{total read/increment/write breakdown of \bigbrain incrementation using 96 processes on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:stacked-96cpus}
\end{figure}

@@

\subsection{20~$\mu$ \bigbrain incrementation}
\subsubsection{25 Processes}

Although tmpfs offers the best performance in terms of makespan duration, Optane
is a close contender. While tmpfs is approximately 5x faster than Optane, Optane can be
as much as 12x faster than on Isilon (Figure~\ref{fig:20mksp25}). By using the results of 
our emulation, we can speculate that Optane will also be significantly faster than 
local disk in a real scenario. The real Optane execution is almost 3x faster than
the fastest local disk emulation configuration (i.e. Memory mode - Spark). As the
emulation results have been found to be faster than the real results, potentially as a side-effect
of increased accessed to cached data, it is expected
that the true makespan of the application running on local disk would be significantly
greater.

The read/increment/write breakdown (Figure~\ref{fig:20stackedp25}) shows that tmpfs
spends the least amount of time executing all the tasks, however, it spends more
time than Optane incrementing the data. Otherwise, tmpfs is nearly 470x faster with
reading and nearly 11x faster at writing than Optane. Despite not reaching memory
speeds, Optane is significantly faster than other persistent storage devices.

\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-real-20bb_25cpus.pdf}
    \caption{real}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-em-20bb_25cpus.pdf}
    \caption{emulated}
\end{subfigure}
\caption{Makespan of the incrementation algorithm processing the 20~$\mu$m BigBrain using
25 processes}\label{fig:20mksp25}
\end{figure*}

\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-real-20bb_25cpus.pdf}
    \caption{real}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-emulated-20bb_25cpus.pdf}
    \caption{emulated}
\end{subfigure}
\caption{Total read/increment/write breakdown of the  incrementation algorithm processing the 20~$\mu$m BigBrain using
25 processes}\label{fig:20stackedp25}
\end{figure*}

\subsubsection{96 Processes}
\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-real-20bb_96cpus.pdf}
    \caption{real}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-em-20bb_96cpus.pdf}
    \caption{emulated}
\end{subfigure}
\caption{Makespan of the incrementation algorithm processing the 20~$\mu$m BigBrain using
96 processes}\label{fig:20mksp96}
\end{figure*}

\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-real-20bb_96cpus.pdf}
    \caption{real}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-emulated-20bb_96cpus.pdf}
    \caption{emulated}
\end{subfigure}
\caption{Total read/increment/write breakdown of the  incrementation algorithm processing the 20~$\mu$m BigBrain using 96 processes}\label{fig:20stackedp96}
\end{figure*}
\subsection{BIDS App Example}
\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bids-app-example/25cores.pdf}
    \caption{25 processes}\label{fig:bm25}
\end{subfigure}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bids-app-example/96cores.pdf}
    \caption{96 processes}\label{fig:bm96}
\end{subfigure}
\caption{Makespan of BIDS App Example 25 and 96 processes on all storage devices. 3 repetitions were performed}
\end{figure*}
